A logical and topological framework for experimental science

### 2018-10-02 Articles/LogicalTopologicalFramework/article.tex
Various changes made as we discussed

\begin{defn}
A \emph{statement} is an object with a truth value, ascertained by a truth function outputting a Boolean, and logical content, defined by satisfying the Consistency Axiom (\ref{consistent}) and the Completeness Axiom (\ref{completeness}). The set of all statements is denoted $\mathcal{S}$. The truth function is denoted $\truth:\mathcal{S}\to\mathbb{B}$.
\end{defn}

@@

This axiom captures the general notion of content of an individual statement. Our earlier example of two statements about electron mass cannot both be true, but all other combinations (both false, one true and one false) are consistent truth assignments. We next define the possibilities of a statement in terms of consistent truth values of single statements.

\begin{defn}
The \emph{possibilities} of a statement $\stmt\in\mathcal{S}$ are the possible truth values allowed by the content of the statement. We define the possibilities function as
$$
\possFn:\mathcal{S}\to\{\{\FALSE,\TRUE\},\{\TRUE\},\{\FALSE\}\}
$$
where $\possFn(\stmt)$ is defined to be the set $S\subseteq\mathbb{B}$ of truth values such that $\{\stmt\},\{B\}$ is a consistent truth assignment for each $B\in S$. 
\end{defn}

@@

The notion of possibilities of a statement captures the difference between statements whose truth values can be obtained a priori (such as mathematical statements or tautologies), and those which must be experimentally tested. It is immediate that $\truth(\stmt)\in\possFn(\stmt)$ for all statements $\stmt$. In fact, for any collection of statements, assigning them their truth values is always a consistent truth assignment, since we assume true scientific statements to be non-contradictory. However, there are many consistent truth assignments which do not arise from the truth values of the included statements. Conversely, for any consistent truth assignment, each truth value $t_{\lambda}$ must be in $\possFn(\stmt_{\lambda})$. We now can define the Completeness Axiom:

\begin{axiom}[Completeness Axiom]
\label{completeness}
Given a collection of statements $\{\stmt_{\lambda}\}_{\lambda\in\Lambda}$ for some (possibly uncountable) index set $\Lambda$, let $A\subseteq\Lambda$, and let $\{t_a\}_{a\in A}$ be a consistent truth assignment for $\{\stmt_a\}_{a\in A}$. Then there exists a consistent truth assignment $\{t_{\lambda}\}_{\lambda\in\Lambda}$ for $\{\stmt_{\lambda}\}_{\lambda\in\Lambda}$ such that $t_{\lambda} = t_a$ for all $\lambda\in A$. 
\end{axiom}
This axiom expands the notion of what is considered to be a possible truth value. No possible truth value of any statement can lead to a paradox in larger collections of statements. These two axioms define our basic collection of statements. Next, we will give an additional axiom which will allow us to create new statements from old ones.

\begin{axiom}[Construction axiom]
Let $f_{\mathbb{B}}$ be a Boolean function $f_{\mathbb{B}}:\mathbb{B}^{\Lambda}\to\mathbb{B}$, where $\Lambda$ is any cardinal. Then there exists a function on the set of statements $f:(\mathcal{S}^*)^{\Lambda}\to\mathcal{S}$ such that
		$$\truth(f(\stmt_1, ..., \stmt_n)) = f_{\mathbb{B}}(\truth(\stmt_1), ..., \truth(\stmt_n))$$
with the same relationship holding for every consistent truth assignment. 
\end{axiom}

The point of the Construction axiom is to extend our usual Boolean logical operations to statements. We will now define them in this context.

@@

It is important to note that the possibilities of a statement composed from other statements cannot be determined just by looking at the truth values and possibilities of individual statements alone. For example, two statements $\stmt_1$ and $\stmt_2$ which both have $\mathbb{B}$ as their possibilities may have any subset of $\mathbb{B}$ as the possibilities of their conjunction $\stmt_1\AND\stmt_2$ (regardless of the truth values).

@@

\begin{defn}
Two statements $\stmt_1$ and $\stmt_2$ are \emph{equivalent}, denoted $\stmt_1 \equiv \stmt_2$, if and only if $$\stmt_0:=(\stmt_1 \AND \stmt_2) \OR (\NOT\stmt_1 \AND \NOT\stmt_2)$$ has $\possFn(\stmt_0)=\{\TRUE\}$. 
\end{defn}

@@

An alternative way to view statement equivalence is as follows:

@@

It is an immediate consequence of the definition that all tautologies are equivalent, and all contradictions are equivalent. We will interchangeably denote tautologies and their equivalence class by $\tautology$, and similarly contradictions and their equivalence class will be denoted by $\contradiction$. Proposition \ref{prop_equivalent_is_iff} allows one to quickly see that if two Boolean functions are equal, then the corresponding functions on $\mathcal{S}$ must give equivalent results. We then arrive at the following:

@@

While $\mathcal{S}$ is not technically a Boolean algebra itself (since e.g. there are many nonequal, but equivalent, tautologies and contradictions), it will enjoy many of the useful consequences of being a Boolean algebra when using statement equivalence rather than equality. We call the operations and properties defined thus far the \emph{algebra of statements}. Many of the methods utilized henceforth will be based on these tools.

@@

We can now exhibit some lattice-type properties of $\mathcal{S}$.

\begin{prop}
	Every subset $S \subseteq \stmtSet$ of statements has a supremum. That is, there exists an element $\bar{\stmt} \in \stmtSet$ such that $\stmt \narrower \bar{\stmt}$ for all $\stmt \in S$. 
\end{prop}

@@

Next, suppose $\stmt_0\in\mathcal{S}$ also has $\stmt\narrower\stmt_0$ for all $\stmt\in S$. Then by Proposition \ref{prop_narrowness_properties} (vii), it follows that $\bar{\stmt}\narrower\stmt_0$. Thus $\bar{\stmt}$ is our supremum. By antisymmetry of $\narrower$, it is unique up to statement equivalence. 
\end{proof}

\begin{cor}
	Every subset $S \subseteq \stmtSet$ of statements has an infimum. That is, there exists an element $\underline{\stmt} \in \stmtSet$ such that $\underline{\stmt} \narrower \stmt$ for all $\stmt \in S$. 
\end{cor}
\begin{proof}
Let $S\subseteq\mathcal{S}$ be a set of statements. consider the set $\NOT S$ consisting of all negations of elements of $S$. Then let $\stmt_0 = \sup\NOT S$. Then since $\NOT\stmt \narrower \stmt_0$ for all $\stmt\in S$, using Prop \ref{boolean_properties} we have that $\NOT\stmt_0\narrower\stmt$ for all $\stmt\in S$, and similar to the above proof, it is a greatest lower bound. We conclude that $\underline{\stmt} =\NOT\stmt_0$ is the infimum we sought. By antisymmetry of $\narrower$, it is unique up to statement equivalence. 
\end{proof}

It follows that the set of equivalence classes of statements is a \emph{complete} Boolean algebra. The supremum and infimum are constructed in the set of statements, but they are only unique up to statement equivalence.

@@

The collection $\vstmtSet$ of verifiable statements is not required to be closed under negation. To see why, consider for example the statement ``particles $A$ and $B$ are of different mass". If this statement is true, we could arrange for a sequence of progressively more precise measuring equipment to compare them, and eventually we would find out. Now, the negation of this statement, ``particles $A$ and $B$ have precisely the same mass" cannot be verified experimentally, since for any measuring apparatus, we will not know if the difference is just within a margin of error. However, we will be interested in certain subsets of $\vstmtSet$ with stronger closure properties.

\begin{defn}
The set of \emph{decidable statements}, denoted $\dstmtSet$, is the subset of $\vstmtSet$ consisting of statements $\stmt\in\vstmtSet$ such that $\NOT\stmt\in\vstmtSet$ as well. 
\end{defn}

\begin{rem}
Decidable statements may be thought of as those which can be shown to be true or false experimentally, or equivalently, as those pairs consisting of a statement and its negation which can both be verified. Given a pair of statements in $\vstmtSet$ which are negations of each other, we can simply run their associated tests for arbitrarily long (using e.g. a similar diagonalization approach to above) to see which one terminates successfully. 
\end{rem}

\begin{prop}\label{prop_decidable_AND_OR}
The collection $\dstmtSet$ of decidable statements is closed under finite conjunction and disjunction.
\end{prop}
\begin{proof}
	Let $\{\stmt_i\}_{i=1}^{n}$ be a finite collection of decidable statements. Then $$\{\stmt_i\}_{i=1}^{n} \subseteq \vstmtSet\text{ and }\{\NOT\stmt_i\}_{i=1}^{n} \subseteq \vstmtSet$$ are each collections of verifiable sets. Now, consider the conjunction $\bigAND\limits_{i=1}^{n} \stmt_i$. This a verifiable statement by the Closure axiom. Further, the negation $\NOT \bigAND\limits_{i=1}^{n} \stmt_i \equiv \bigOR\limits_{i=1}^{n} \NOT \stmt_i$ is also verifiable by the Closure axiom. We conclude that $\bigAND\limits_{i=1}^{n} \stmt_i\in\dstmtSet$.

	Similarly, the disjunction $\bigOR\limits_{i=1}^{n} \stmt_i$ and its negation $\NOT \bigOR\limits_{i=1}^{n} \stmt_i \equiv \bigAND\limits_{i=1}^{n} \NOT \stmt_i$ are verifiable statements by the closure axioms.
\end{proof}

@@

We will generally denote possibilities by $x$, the set of all possibilities for an experimental domain by $X$. Since possibilities are statements, we use the same equivalence relation and ordering. The set of equivalence classes in $X$ will be denoted by $X^*$. We will often use the phrasing, ``a possibility $x$ \emph{determines} the truth value of a statement". This follows from the fact that if a possibility is true (or assigned true in a consistent assignment), then all statements incompatible must be assigned false, and all statements compatible are necessarily broader and hence must be assigned true. We now show a few of the fundamental properties of possibilities.

@@

\begin{prop}\label{exactly_one_poss}
In the set of possibilities $X$ for any experimental domain $\edomain$, exactly one equivalence class of possibilities is true. 
\end{prop}
\begin{proof}
First, we show at most one possibility is true. Suppose $x_1,x_2\in X$ are both true. Since they are true simultaneously, they are not incompatible. Then we have $x_1\narrower x_2$ and $x_2\narrower x_1$, and so $x_1\equiv x_2$.

@@

With the tools we have thus far developed, we are now ready to answer a fundamental question: what is the size of the largest set amongst which scientific methods can distinguish? This is answered by looking at the possible cardinalities of sets of (equivalence classes of) possibilities. The requirement of a countable basis for an experimental domain is crucial. Recall Theorem \ref{cardinality_thm}, which states that $X^*$ has cardinality at most that of the continuum.

@@

We are now ready to define the topology on $X^*$, the set of equivalence classes of possibilities.

\begin{prop}[Theorem \ref{main}, the topology]
\label{topology}
The collection of verifiable sets $\mathsf{T}_X=\{U(\stmt)|\stmt\in\edomain\}$ for $\stmt\in\edomain$ forms a topology on the set $X^*$. 
\end{prop}
\begin{proof}
First, recall that $\tautology$ and $\contradiction$ are verifiable sets. We have $U(\tautology) = X$ and $U(\contradiction) = \emptyset$.

@@

\begin{cor}[Theorem \ref{main}, countability]
The collection $U(\basis)\cup\{X\}$ of verifiable sets for a basis of an experimental domain, together with the set of all possibilities, forms a sub-basis for the topology of experimental possibility $T_X$.
\end{cor}
\begin{proof}
It is clear that $U(\basis)$ can generate all nontrivial open sets by union (countable, and hence arbitrary, disjunction) and finite intersection (conjunction). Because $\tautology$ is not necessarily in $\basis$, we need to add $\{X^*\}$ to the collection to reach the residual possibility, if it exists. 
\end{proof}

@@

\begin{cor}[Theorem \ref{main}, countability]
The topology of experimental possibility is second-countable.
\end{cor}
\begin{proof}
The collection $U(\basis)\cup\{X^*\}$ is a countable sub-basis which may be extended to a countable basis by including finite intersections. 
\end{proof}

@@

\begin{prop}[Theorem \ref{main}, $\mathsf{T}_0$ separability]
The topology of experimental possibility has $\mathsf{T}_0$ separability for any experimental domain. 
\end{prop}
\begin{proof}
Let $x_1,x_2\in X^*$. It will suffice to show that there exists a verifiable statement $\mathsf{e}\in\basis$ whose verifiable set contains one but not the other. Because possibilities may be expressed as minterms of basis elements, if $x_1$ and $x_2$ are distinct, then there exists a basis element $b\in\basis$ which is negated in, without loss of generality, $x_1$ but not $x_2$. Then $x_2\in U(b)$ but $x_1\notin U(b)$, as required. 
\end{proof}

>>>>

We begin the axiomatization of our universe of scientific statements by declaring existence of our collection. Following this, the special properties of its elements are built up.

\begin{axiom}[Existence Axiom]
There exists a collection $\mathcal{S}$ of \emph{statements} whose properties are as follows. Each $\stmt\in\mathcal{S}$ is an object with a truth value, ascertained by a truth function outputting a Boolean, and logical content, defined by satisfying the Consistency Axiom (\ref{consistent}) and the Completeness Axiom (\ref{completeness}). The set of all statements is denoted $\mathcal{S}$. The truth function is denoted $\truth:\mathcal{S}\to\mathbb{B}$.
\end{axiom}

@@

This axiom captures the general notion of content of an individual statement. Our earlier example of two statements about electron mass cannot both be true, but all other combinations (both false, one true and one false) are consistent truth assignments. 
We now can define the Completeness Axiom:

\begin{axiom}[Completeness Axiom]
\label{completeness}
Given a collection of statements $\{\stmt_{\lambda}\}_{\lambda\in\Lambda}$ for some (possibly uncountable) index set $\Lambda$, let $A\subseteq\Lambda$, and let $\{t_a\}_{a\in A}$ be a consistent truth assignment for $\{\stmt_a\}_{a\in A}$. Then there exists a consistent truth assignment $\{t_{\lambda}\}_{\lambda\in\Lambda}$ for $\{\stmt_{\lambda}\}_{\lambda\in\Lambda}$ such that $t_{\lambda} = t_a$ for all $\lambda\in A$. 
\end{axiom}
This axiom restricts the notion of what is considered to be a consistent truth assignment. No consistent truth assignment for any single statement can lead to a paradox in larger collections of statements. We next define the possibilities of a statement in terms of consistent truth values of single statements.

\begin{defn}
The \emph{possibilities} of a statement $\stmt\in\mathcal{S}$ are the possible truth values allowed by the content of the statement. We define the possibilities function as
$$
\possFn:\mathcal{S}\to\{\{\FALSE,\TRUE\},\{\TRUE\},\{\FALSE\}\}
$$
where $\possFn(\stmt)$ is defined to be the set $B\subseteq\mathbb{B}$ of truth values such that $\{\stmt\},\{t\}$ is a consistent truth assignment for each $t\in B$. 
\end{defn}

@@

The notion of possibilities of a statement captures the difference between statements whose truth values can be obtained a priori (such as mathematical statements or tautologies), and those which must be experimentally tested. For any collection of statements, assigning them their truth values is always a consistent truth assignment, since we assume true scientific statements to be non-contradictory. However, there are many consistent truth assignments which do not arise from the truth values of the included statements. Conversely, for any consistent truth assignment, each truth value $t_{\lambda}$ must be in $\possFn(\stmt_{\lambda})$.

It is important to note that the possibilities of a statement composed from other statements cannot be determined just by looking at the truth values and possibilities of individual statements alone. For example, two statements $\stmt_1$ and $\stmt_2$ which both have $\mathbb{B}$ as their possibilities may have any subset of $\mathbb{B}$ as the possibilities of their conjunction $\stmt_1\AND\stmt_2$ (regardless of the truth values). Next, we will give an additional axiom which will allow us to create new statements from old ones.

\begin{axiom}[Construction axiom]
Let $f_{\mathbb{B}}$ be a Boolean function $f_{\mathbb{B}}:\mathbb{B}^{\Lambda}\to\mathbb{B}$, where $\Lambda$ is an index set. Then there exists a function on the set of statements $f:\mathcal{S}^{\Lambda}\to\mathcal{S}$ with the following requirements. First, the truth value is determined by the Boolean function:
\begin{equation}
\label{constructionrequirement}
\truth(f(\{\stmt_{\lambda}\}_{\lambda\in\Lambda}) = f_{\mathbb{B}}(\{\truth(\stmt_{\lambda})\}_{\lambda\in\Lambda}).
\end{equation}
Second, any consistent truth assignment for a collection of statements $S$ which includes the statement $f(\{\stmt_{\lambda}\}_{\lambda\in\Lambda})$ extends to a consistent truth assignment for $S\cup\{\stmt_{\lambda}\}_{\lambda\in\Lambda}$ such that the assignment to $f(\{\stmt_{\lambda}\}_{\lambda\in\Lambda})$ is given by $f_{\mathbb{B}}(\{t_{\lambda}\}_{\lambda\in\Lambda}$. 
\end{axiom}

\begin{rem}
Note that in the second condition of the axiom, this means that if $S$ already contains some of the $\stmt_{\lambda}$'s, then their truth assignments must already be compatible in the sense of the second condition above. 
\end{rem}

The Construction axiom allows us to turn any Boolean function into a function on the collection of statements. In particular, the usual logic operations can be applied to $\mathcal{S}$.

@@

@@

\begin{defn}
Two statements $\stmt_1$ and $\stmt_2$ are \emph{equivalent}, denoted $\stmt_1 \equiv \stmt_2$, if and only if in all consistent truth assignments, $\stmt_1$ and $\stmt_2$ must be assigned the same truth value. 
\end{defn}

@@

It is an immediate consequence of the definition that all tautologies are equivalent, and all contradictions are equivalent. We will interchangeably denote tautologies and their equivalence class by $\tautology$, and similarly contradictions and their equivalence class will be denoted by $\contradiction$. An alternative way to view statement equivalence is as follows:

@@

Proposition \ref{prop_equivalent_is_iff} allows one to quickly see that if two Boolean functions are equal, then the corresponding functions on $\mathcal{S}$ must give equivalent results. We then arrive at the following:

@@

Note that $\mathcal{S}$ is not a Boolean algebra itself (since e.g. there are many nonequal, but equivalent, tautologies and contradictions). Henceforth, unless otherwise noted, \emph{we will exclusively use the notation $\mathcal{S}$ for the collection of equivalence classes of statements}. The Boolean operations with $\tautology$ and $\contradiction$ (now each individual equivalence classes) endow the set of equivalence classes $\mathcal{S}$ with the structure of a Boolean algebra. Further, it is a \emph{complete Boolean algebra}, since the usual operations of conjunction and disjunction allow one to obtain an $\inf$ and $\sup$ for any subset $S\subseteq\mathcal{S}$.

We call the operations and properties defined thus far the \emph{algebra of statements}. Many of the methods utilized henceforth will be based on these tools.

@@

We can now exhibit some lattice-type properties of $\mathcal{S}$ endowed with $\narrower$.

\begin{prop}
The supremum (respectively infimum) of any set $S \subseteq \stmtSet$ of statements with respect to $\narrower$ is equal to the Boolean supremum (respectively infimum). That is, for any collection $S$, the conjunction $\OR S$ (respectively $\AND S$ is the supremum (respectively infimum) with respect to $\narrower$. Denote them by $\OR S := \bar{\stmt} \in \stmtSet$ and $\AND S := \underline{\stmt}\in\stmtSet$. 
\end{prop}

@@

Next, suppose $\stmt_0\in\mathcal{S}$ also has $\stmt\narrower\stmt_0$ for all $\stmt\in S$. Then by Proposition \ref{prop_narrowness_properties} (vii), it follows that $\bar{\stmt}\narrower\stmt_0$. Thus $\bar{\stmt}$ is our supremum. By antisymmetry of $\narrower$, it is unique up to statement equivalence.

For the infimum, consider the set $\NOT S$ consisting of all negations of elements of $S$. Then let $\stmt_0 = \sup\NOT S$. Then since $\NOT\stmt \narrower \stmt_0$ for all $\stmt\in S$, using Prop \ref{boolean_properties} we have that $\NOT\stmt_0\narrower\stmt$ for all $\stmt\in S$, and similar to the above proof, it is a greatest lower bound. We conclude that $\underline{\stmt} =\NOT\stmt_0$ is the infimum we sought. By antisymmetry of $\narrower$, it is unique up to statement equivalence. 
\end{proof}

@@

The collection $\vstmtSet$ of verifiable statements is not required to be closed under negation. To see why, consider for example the statement ``particles $A$ and $B$ are of different mass". If this statement is true, we could arrange for a sequence of progressively more precise measuring equipment to compare them, and eventually we would find out. Now, the negation of this statement, ``particles $A$ and $B$ have precisely the same mass" cannot be verified experimentally, since for any measuring apparatus, we will not know if the difference is just within a margin of error.

@@

We will generally denote possibilities by $x$, the set of all possibilities for an experimental domain by $X$. Since possibilities are statements, we can use the same ordering. We will often use the phrasing, ``a possibility $x$ \emph{determines} the truth value of a statement". This follows from the fact that if a possibility is true (or assigned true in a consistent assignment), then all statements incompatible must be assigned false, and all statements compatible are necessarily broader and hence must be assigned true. We now show a few of the fundamental properties of possibilities.

@@

\begin{prop}\label{exactly_one_poss}
In the set of possibilities $X$ for any experimental domain $\edomain$, exactly one of the possibilities is true. 
\end{prop}
\begin{proof}
First, we show at most one possibility is true. Suppose $x_1,x_2\in X$ are both true. Since they are true simultaneously, they are not incompatible. Then we have $x_1\narrower x_2$ and $x_2\narrower x_1$, and so $x_1\equiv x_2$.

@@

With the tools we have thus far developed, we are now ready to answer a fundamental question: what is the size of the largest set amongst which scientific methods can distinguish? This is answered by looking at the possible cardinalities of sets of possibilities. The requirement of a countable basis for an experimental domain is crucial. Recall Theorem \ref{cardinality_thm}, which states that $X$ has cardinality at most that of the continuum.

@@

We are now ready to define the topology on $X$, the set of possibilities.

\begin{prop}[Theorem \ref{main}, the topology]
\label{topology}
The collection of verifiable sets $\mathsf{T}_X=\{U(\stmt)|\stmt\in\edomain\}$ for $\stmt\in\edomain$ forms a topology on the set $X$. 
\end{prop}
\begin{proof}
First, recall that $\tautology$ and $\contradiction$ are verifiable sets. We have $U(\tautology) = X$ and $U(\contradiction) = \emptyset$.

@@

\begin{cor}[Theorem \ref{main}, countability]
The collection $U(\basis)\cup\{X\}$ of verifiable sets for a basis of an experimental domain, together with the set of all possibilities, forms a sub-basis for the topology of experimental possibility $T_X$.
\end{cor}
\begin{proof}
It is clear that $U(\basis)$ can generate all nontrivial open sets by union (countable, and hence arbitrary, disjunction) and finite intersection (conjunction). Because $\tautology$ is not necessarily in $\basis$, we need to add $\{X\}$ to the collection to reach the residual possibility, if it exists. 
\end{proof}

@@

\begin{cor}[Theorem \ref{main}, countability]
The topology of experimental possibility is second-countable.
\end{cor}
\begin{proof}
The collection $U(\basis)\cup\{X\}$ is a countable sub-basis which may be extended to a countable basis by including finite intersections. 
\end{proof}

@@

\begin{prop}[Theorem \ref{main}, $\mathsf{T}_0$ separability]
The topology of experimental possibility has $\mathsf{T}_0$ separability for any experimental domain. 
\end{prop}
\begin{proof}
Let $x_1,x_2\in X$. It will suffice to show that there exists a verifiable statement $\mathsf{e}\in\basis$ whose verifiable set contains one but not the other. Because possibilities may be expressed as minterms of basis elements, if $x_1$ and $x_2$ are distinct, then there exists a basis element $b\in\basis$ which is negated in, without loss of generality, $x_1$ but not $x_2$. Then $x_2\in U(b)$ but $x_1\notin U(b)$, as required. 
\end{proof}
